\LARGE \bf Planning with Attitude

### 2020-10-09 paper/main.tex

\author{Brian Jackson$^1$ and Zachary Manchester$^1$%
    \thanks{
        $^1$Robotics Institute, 
        Carnegie Mellon University, 
        5000 Forbes Ave, Pittsburgh, PA, USA
    }
}

@@

\begin{abstract}
Planning and controlling trajectories for floating-base robotic systems that experience
large attitude changes is challenging due to the nontrivial group structure of 3D
rotations. This paper introduces an accessible and unified approach for tracking control
and optimization-based planning on the space of rotations based on vector calculus and
linear algebra. The methodology is used to derive an extension of the Linear-Quadratic
Regulator (LQR) to systems with arbitrary attitudes, which we call Multiplicative LQR
(MLQR). We compare MLQR to a state-of-the-art geometric tracking controller designed
specifically for quadrotors and demonstrate superior tracking and robustness. An
iterative variant of MLQR is compared to existing methods on several constrained
trajectory optimization problems, again demonstrating improved runtime performance. 
\keywords{
    motion planning and control, quaternions, optimal control, linear quadratic regulator
}
\end{abstract}

@@

    In this paper, we make a departure from previous approaches to geometric planning and 
    control that rely heavily on ideas and notation from differential geometry, 
    and instead use only basic mathematical tools from linear algebra and calculus that 
    should be familiar to most roboticists. 
    Similar to \cite{Mandic2011,Xu2016}, in Sec. \ref{sec:Quaternion_Calculus} we introduce 
    a quaternion differential calculus, but take a significantly simpler and more general 
    approach that unifies both planning and control, enabling straight-forward adaptation of 
    existing algorithms to systems with quaternion states. 
    To make this concrete, in Sec. \ref{sec:MLQR} we apply our approach to derive an 
    extension to the linear-quadratic regulator (LQR) that we call multiplicative LQR (MLQR), 
    which is the control dual to the multiplicative extended Kalman filter (MEKF) from the 
    state estimation literature \cite{markley2014fundamentals}. 
    In Sec. \ref{sec:experiments} we provide several simulation results demonstrating the 
    application of our method to tracking control and constrained trajectory optimization.
    In summary, our contributions include:

    \begin{itemize}
        \item A unified approach to quaternion differential calculus entirely based on standard vector calculus and linear algebraic operations
        \item Derivation of multiplicative LQR, an adaptation of LQR to the control of systems with quaternion states
        \item A set of benchmark motion-planning and trajectory-tracking problems in which we compare our approach to existing methods commonly used in the robotics community to account for 3D rotations
    \end{itemize}

@@

        We will sometimes find it helpful to create a quaternion with zero scalar part from 
        a vector $r \in \R^3$. We denote this operation as,
        \begin{equation}
            \hat{r} = H r \equiv \begin{bmatrix} 0 \\ I_3 \end{bmatrix} r.
        \end{equation}
        Unit quaternions rotate a vector through the operation 
        $\hat{r}' = \q \otimes \hat{r} \otimes \q^*$. 
        This can be equivalently expressed using matrix multiplication as
        \begin{align} 
            r' &= H^T L(q) R(q)^T H r = A(q)r , \label{eq:quaternion_rotation}
        \end{align}
        where $A(q)$ is the rotation matrix in terms of the elements of the quaternion 
        \cite{markley2014fundamentals}.

    \subsection{Rigid Body Dynamics} \label{sec:rigidbody_dynamics}
        In the current work we will restrict our focus to rigid bodies moving freely in 3D 
        space. That is, we consider systems with dynamics of the following form:
        \begin{equation} \label{eq:rigid_body_dynamics}
            x = \begin{bmatrix} r \\ R \\ v \\ \omega \end{bmatrix}, \quad 
            \dot{x} = \begin{bmatrix} 
                v \\ 
                \texttt{kinematics}(R,\omega) \\ 
                \frac{1}{m} F_G(x,u) \\ 
                J^{-1}(\tau_L(x,u) - \omega \times J \omega) 
            \end{bmatrix}
        \end{equation}
        where $x$ and $u$ are the state and control vectors, $r \in \R^3$ is the position, 
        $R \in SO(3)$ is the attitude, $v \in \R^3$ is the linear velocity, and 
        $\omega \in \R^3$ is the angular velocity. $m \in \R$ is the mass, 
        $J \in \R^{3\times3}$ is the inertia matrix, $F_G(x,u) \in \R^3$ are the forces in the 
        global frame, $\tau_L(x,u)$ are the moments in the local (body) frame, and 
        $\texttt{kinematics}$ are the kinematics for the chosen attitude representation. The 
        kinematics for unit quaternions are
        \begin{equation}
            \dot{q} = \half \q \otimes \hat{\omega} = \half L(q) H \omega.
        \end{equation}

@@

\section{Multiplicative LQR} \label{sec:MLQR}
    Leveraging the methods from the previous section, we derive multiplicative LQR
    (MLQR), a variant of LQR that correctly accounts for the group structure of
    rotations. For concreteness, we consider a system with rigid body dynamics, as
    presented in Sec. \ref{sec:rigidbody_dynamics}, and design a controller to stabilize
    the system about a dynamically feasible discretized reference trajectory
    $\bar{x}_{0:N}, \bar{u}_{0:N}$ with $N$ time steps.
    We begin by linearizing the dynamics about the reference trajectory using \eqref{eq:quat_jacobian}. Our linearized error dynamics become
    \begin{equation} \label{eq:linearized_dynamics}
        \delta x_{k+1} = A_k \delta x_k + B_k \delta u_k 
    \end{equation}
    where \begin{equation}
        \begin{aligned}
            A_k = E(\bar{x}_{k+1})^T \pdv{f}{x}|_{\bar{x}_k,\bar{u}_k} E(\bar{x}_k), \\
            B_k = E(\bar{x}_{k+1})^T \pdv{f}{u}|_{\bar{x}_k,\bar{u}_k},
        \end{aligned}
    \end{equation}
    and $\delta x_k \in \R^{12}$ and $E(x_k) \in \R^{12 \times 13}$ are the state error and state error Jacobian:
    \begin{equation} \label{eq:state_error}
        \setlength\arraycolsep{1pt}
        \delta x_k = \begin{bmatrix} 
            r_k - \bar{r}_k \\ \varphi^{-1}(\bar{\q}_k^{-1} \otimes \q_k) \\ v_k - \bar{v}_k \\ \omega_k - \bar{\omega}_k 
        \end{bmatrix}\!, \;
        E(x) = \begin{bmatrix}
            I_3 & & & \\
            & G(q) & & \\
            & & I_3 & \\
            & & & I_3 \\
        \end{bmatrix}\!.
    \end{equation}

    With our linearized system, we can apply the traditional LQR Riccati recursion,
    \begin{multline}
         P_k = W_k +  A_k^T P_{k+1} A_k \\
         - A_k^T P_{k+1}^T B_k (R_k + B_k^T P_{k-1} B_k)^{-1} B_k^T P_{k+1} A_k ,
    \end{multline}
    which minimizes the quadratic cost function
    \begin{equation}
        \half \sum_{k=0}^N \delta x_k^T Q_k \delta x_k + \delta u_k^T R_k \delta u_k
    \end{equation}
    of the state \emph{error}, where $Q_k \in \R^{12 \times 12}$ is the weight matrix on
    the state \emph{error}, and $R_k \in \R^m$ is the weight matrix on the controls. From
    this, we get our non-linear feedback policy
    \begin{equation} \label{eq:mlqr_control}
        u = K_k \delta x_k + \bar{u}_k.
    \end{equation}
    where $\delta x_k$ is computed online using the non-linear error function
    \eqref{eq:state_error} and the linear feedback gain $K_k$ is calculated using the
    standard LQR formula:
    \begin{equation} \label{eq:LQR_gain}
         K_k := -(R_k + B_k^T P_{k+1} B_k)^{-1} B_k^T P_{k+1} A_k.
    \end{equation}

    The multiplicative Linear Quadratic Regulator is summarized as follows:
    \begin{enumerate}
        \item Pick $Q_k \in \R^{12 \times 12}$ to weight the state \textit{error}, and $R_k \in \R^{m \times m}$
        \item Linearize the dynamics using the state error Jacobian \eqref{eq:state_error}
        \item Compute $K_k$ using the standard LQR Riccati recursion
        \item Online, compute the control using \eqref{eq:state_error} and \eqref{eq:mlqr_control}.
    \end{enumerate}

    \subsection{Constrained Iterative MLQR}
        As demonstrated above, it is very straight-forward to adapt LQR to use
        quaternions. We similarly use this technique to adapt nonlinear trajectory
        optimization algorithms. Here we present the modifications to the ALTRO solver
        \cite{howell2019altro}, which uses iterative LQR (iLQR) combined with an
        augmented Lagrangian approach to handle constraints.

        For each iteration of iLQR within ALTRO, we can calculate a quadratic
        approximation of the cost function using \eqref{eq:quat_gradient} and
        \eqref{eq:quat_hessian}. Identical to \eqref{eq:linearized_dynamics}, we use
        \eqref{eq:quat_jacobian} to linearize the dynamics and use MLQR to solve for the
        feedback and feedforward gains.

        The only other modification to the algorithm is during the iLQR ``forward pass''
        that simulates the system with the closed-loop MLQR controller, where we use
        \eqref{eq:state_error} to calculate the nonlinear feedback policy during the
        forward roll-outs.
        For robustness, we may also add an additional criterion to the line search that
        checks for singularities in the three-parameter error state which, for the Cayley
        map, occur at \ang{180}, at which point the step size should be reduced.

    \subsection{Quaternion Cost Functions} \label{sec:cost_functions}
        In addition to the straight-forward modifications to the LQR algorithm itself, we
        need to carefully consider the types of cost functions we minimize. We frequently
        minimize costs that penalize distance from a goal state, e.g. $\half (x-x_g)^T Q
        (x-x_g)$; however, na\"ive substraction of unit quaternions is ill-defined. We
        propose two different cost functions that accomplish similar behavior.
        For sake of clarity and space, we only consider the costs on the quaternion
        variables: costs on the other states and the control variables remain unaffected.

@@

        \begin{figure}
            \centering
            \includegraphics[height=4cm, width=8cm]{figures/timing_chart.tikz}
            \caption{Timing results for the airplane barrell roll, quadrotor
                flip, and quadrotor zig-zag examples using iLQR with
                roll-pitch-yaw Euler angles (RPY), quaternions, and iMLQR. The
                timing result for the Quadrotor flip with Euler angles is omitted
                since it failed to converge.
            }
            \label{fig:timing_chart}
        \end{figure}

        \subsubsection{Quadrotor Zig-Zag}
            \begin{figure}[t]
                \centering
                \includegraphics[height=5.2cm]{figures/zig_zag_tracking.png}
                \caption{Zig-zag trajectory. The black quadrotors indicate
                    the reference trajectory generated by iMLQR. The yellow
                    quadrotors track the reference trajectory using time-varying
                    MLQR, while the red uses the HFCA controller from
                    \cite{watterson2020control}.
                }
                \label{fig:zig-zag}
            \end{figure}    
            \begin{figure}[t]
                \centering
                \includegraphics[height=5.2cm,trim={0 3cm 0 0},clip]{figures/quad_flip.png}
                \caption{Snapshots of the quadrotor flip trajectory. The
                    red-colored quadrotors represent the state near t=0 s and the
                    green-colored quadrotors represent the state near t=5.0 s
                }
                \label{fig:quad_flip}
            \end{figure}

        We set up a fairly simple trajectory optimization problem to achieve a
        ``zig-zag'' pattern, as shown in Fig. \ref{fig:zig-zag} using cost funtions of
        the form
        \begin{multline}
                  \half \sum_{k \in \mathcal{N}} \ell(x_k, x_\text{nom}, Q_\text{nom}) 
                + \half \sum_{k \in \mathcal{W}} \ell(x_k, \bar{x}_k, Q_w) \\
                + \half \sum_{k=1} u_k^T R u_k
        \end{multline}
        with $\mathcal{W} = \{33,66,101\}$, $\mathcal{N} = \{1:101\} \setminus
        \mathcal{W}$, and the waypoint positions were $(10,0,1)$,$(-10,0,1)$ and
        $(0,10,0)$. We compared the two cost functions from Sec.
        \ref{sec:cost_functions}, using a standard quadratic penalty on the
        non-quaternion states.
        \begin{align}
            \ell_\text{err}(x, \bar{x}, Q) &= \delta x^T Q \delta x \\ 
            \ell_\text{geo}(x, \bar{x}, Q) &= (x-\bar{x})^T \bar{Q} (x-\bar{x}) + w (1
                \pm \bar{q}^T q)
        \end{align}
        where $\bar{Q} = \text{diag}(Q_r, \vec{0}_4, Q_v, Q_\omega)$, where
        $Q_r,Q_v,Q_\omega$ are the weights of $Q$ for position, and linear and angular
        velocity.
        We used the following weighting matrices: 
        $Q_\text{nom} = \text{diag}(10^{-5} I_6,10^{-3} I_6)$, 
        $Q_w = \text{diag}(10^3 I_3,I_9)$, 
        $Q_N = \text{diag}(10 I_3, 100 I_3, 10 I_6)$.
        For the geodesic weights $w$, we used 0, 1, and 10 for the nominal, waypoint, and
        terminal cost functions, respectively.

        The geodesic cost function performed significantly better than the error
        quadratic, converging in 10 ms and 13 iterations, compared to 53 ms in 29
        iterations. In addition to better convergence behavior, the geodesic cost is
        about half as expensive to compute, which is unsurprising when comparing the
        gradients and Hessians in Sec. \ref{sec:cost_functions}.

@@

@@

	    \begin{figure}[h]
	        \centering
	        \includegraphics[width=\columnwidth,height=4.5cm]{figures/tracking_err.tikz}
	        \caption{Tracking error over time for the zig-zag trajectory using MLQR and the HFCA geometric controller. Band-limited white noise wrenches of variance $w$ are added to the continuous dynamics. }
	        \label{fig:tracking_err}
	    \end{figure}
        \vspace{-20pt}

	\subsection{Tracking Control}
	    This section provides examples using MLQR for tracking and stabilizing control
	    for a quadrotor. We compare our MLQR controller to HFCA, a state-of-the-art
	    geometric tracking controller \cite{watterson2020control} based on the Hopf
	    Fibration.

	    \subsubsection{MLQR Stabilization}
	    To test robustness and the domain of attraction of the MLQR controller, a
	    quadrotor is linearized about hover at the origin and initialized from states
	    uniformly sampled in $[-1,1]$ for position, $[-5,5]$ for linear and angular
	    velocities, and uniformly over the entire space of rotations. The success rates
	    for 2000 trials are given in Table \ref{tab:mlqr_success}, where success is
	    defined as the terminal error (calculated using \eqref{eq:state_error}) being
	    within a unit ball of radius 0.1 after 10.0 seconds. To further analyze the
	    basins of attraction, we uniformly sampled over only roll (which goes singular at
	    \ang{90} for our Euler angle representation), and yaw. The results are shown in
	    Fig. \ref{fig:mlqr_basin}, where the plotted contours are the convex hull of the
	    successfully stabilized initial conditions.

	    MLQR slightly exceeds the robustness of the HFCA geometric controller, while also
	    being more generally applicable, easier to tune, and easier to implement. It is
	    also noteworthy that na\"ively using LQR with quaternions (incorrectly treating
	    them as vectors) results in poorer performance than Euler angles, clearly
	    demonstrating the value of the methodology presented in the current paper.

	    \subsubsection{MLQR Tracking}
	    Time-varying MLQR is used to track the zig-zag trajectory generated in the
	    previous section. %The controller calculated the gains at $0.01 s$ intervals, and
	    used zero-order-hold between intervals. To test tracking performance, zero-mean
	    Gaussian noise was added as force and moments applied to the quadrotor.

	    The trajectories and tracking error for MLQR and HFCA are shown in Figs.
	    \ref{fig:zig-zag}, and \ref{fig:tracking_err}. While the MLQR controller
	    dramatically out-performs HFCA, we found no significant difference in performance
	    between MLQR and LQR using either roll-pitch-yaw Euler angles or a quaternion.
	    Since the tracking controller only deals with relatively small state and attitude
	    errors, this is unsurprising. Additionally, we found the MLQR controller ran
        about 20x faster than the HFCA controller.

	    \begin{figure*}[ht]
	        \begin{minipage}{0.60\linewidth}
                \centering
                \includegraphics[height=6.6cm,width=\textwidth]{figures/mlqr_basin.tikz}
                \caption{Basins of attraction of 4 stabilizing controllers for yaw and roll, calculated from 2000 sample points (shown in black). All points within each contour were successfully stabilized by the controller.}
                \label{fig:mlqr_basin}
            \end{minipage}
	        \hspace{0.05\linewidth}
            \begin{minipage}{0.34\linewidth}
                \centering
                \begin{tabular}{cc}
                    \toprule
                    \textbf{Method} & \textbf{Success} \\
                    \midrule
                    Quat & 61\% \\
                    RPY  & 80\% \\
                    HFCA & 97\% \\
                    MLQR & 99\% \\
                    \bottomrule
                \end{tabular}
                \caption{Success rate for Monte-Carlo analysis of quadrotor stabilization for 2000 initial conditions uniformly over position, orientation, and linear and angular velocities.}
                \label{tab:mlqr_success}
            \end{minipage}
        \end{figure*}

>>>>

\author{Brian Jackson$^1$, Kevin Tracy$^1$, and Zachary Manchester$^1$%
    \thanks{
        $^1$Robotics Institute, 
        Carnegie Mellon University, 
        5000 Forbes Ave, Pittsburgh, PA, USA
    }
}

@@

\begin{abstract}
Planning and controlling trajectories for floating-base robotic systems that experience
large attitude changes is challenging due to the nontrivial group structure of 3D
rotations. This paper introduces an accessible approach for 
optimization-based planning on the space of rotations based on vector calculus and
linear algebra. 
\keywords{
    motion planning and control, quaternions, optimal control, linear quadratic regulator
}
\end{abstract}

@@

    In this paper, we make a departure from previous approaches to geometric planning and 
    control that rely heavily on ideas and notation from differential geometry, 
    and instead use only basic mathematical tools from linear algebra and calculus that 
    should be familiar to most roboticists. 
    Similar to \cite{Mandic2011,Xu2016}, in Sec. \ref{sec:Quaternion_Calculus} we introduce 
    a quaternion differential calculus, but take a significantly simpler and more general 
    approach, enabling straight-forward adaptation of 
    existing algorithms to systems with quaternion states. 
    To make this concrete, in Sec. \ref{sec:Wahbas} we apply this method to the canonical
    Wahbas problem and demonstrate dramatically superior convergence to approaches that
    fail to properly account for the group structure. 
    In Sec. \ref{sec:trajopt} we extend these ideas to the problem of trajectory optimization,
    and detail modifications to ALTRO, a state-of-the-art constrained trajectory optimization
    solver, and demonstrate the performance gains on several constrained benchmark problems.
    To our knowledge, there does not currently exist a solver that is capable of leveraging
    the unique Markovian structure of the fixed-horizon trajectory optimization problem while
    correctly accounting for the group structure of 3D rotations.
    In summary, our contributions include:

    \begin{itemize}
        \item A unified approach to quaternion differential calculus entirely based on
        standard vector calculus and linear algebraic operations control of systems with
        quaternion states
        \item A fast and efficient solver for trajectory optimization problems with 
        attitude dynamics and nonlinear constraints that correctly accounts for the group
        structure of 3D rotations during the solve
    \end{itemize}

@@

        We will sometimes find it helpful to create a quaternion with zero scalar part from 
        a vector $r \in \R^3$. We denote this operation as,
        \begin{equation}
            \hat{r} = H r \equiv \begin{bmatrix} 0 \\ I_3 \end{bmatrix} r.
        \end{equation}
        Unit quaternions rotate a vector through the operation 
        $\hat{r}' = \q \otimes \hat{r} \otimes \q^*$. 
        This can be equivalently expressed using matrix multiplication as
        \begin{align} 
            r' &= H^T L(q) R(q)^T H r = A(q)r , \label{eq:quaternion_rotation}
        \end{align}
        where $A(q)$ is the rotation matrix in terms of the elements of the quaternion 
        \cite{kane1983spacecraftdynamics}.

    \subsection{Rigid Body Dynamics} \label{sec:rigidbody_dynamics}
        In the current work we will restrict our focus to rigid bodies moving freely in 3D 
        space. That is, we consider systems with dynamics of the following form:
        \begin{equation} \label{eq:rigid_body_dynamics}
            x = \begin{bmatrix} r \\ R \\ v \\ \omega \end{bmatrix}, \quad 
            \dot{x} = \begin{bmatrix} 
                v \\ 
                \half \q \otimes \hat{\omega} = \half L(q) H \omega \\ 
                \frac{1}{m} F_G(x,u) \\ 
                J^{-1}(\tau_L(x,u) - \omega \times J \omega) 
            \end{bmatrix}
        \end{equation}
        where $x$ and $u$ are the state and control vectors, $r \in \R^3$ is the position, 
        $R \in SO(3)$ is the attitude, $v \in \R^3$ is the linear velocity, and 
        $\omega \in \R^3$ is the angular velocity. $m \in \R$ is the mass, 
        $J \in \R^{3\times3}$ is the inertia matrix, $F_G(x,u) \in \R^3$ are the forces in the 
        global frame, and $\tau_L(x,u)$ are the moments in the local (body) frame.

@@

\section{Modifying Newton's Method} \label{sec:Wahbas}
    Brief background and description of the problem.

    \subsection{Methodology}
    Brief description of using Newton's method to solve the batch version of Wahbas problem.

    \subsection{Results}

\section{Trajectory Optimization on $\R^n \times SO(3)$}
    Here we outline the modifications to the ALTRO solver \cite{howell2019altro}, to
    solve trajectory optimization problems for rigid bodies, which extends easily to
    arbitrary systems in $R^n \times SO(3)$. ALTRO is an efficient solver for constrained
    nonlinear optimization problems that uses iterative LQR (iLQR) with an augmented
    Lagrangian framework.

    We consider a trajectory optimization problem of the form:
    \begin{mini}[2]
        {x_{0:N},u_{0:N-1}}{\ell_f(x_N) + \sum_{k=0}^{N-1} \ell(x_k,u_k) }{}{}
        \addConstraint{x_{k+1} = f(x_k,u_k)}
        \addConstraint{g_k(x_k,u_k)}{\leq 0}
        \addConstraint{h_k(x_k,u_k)}{=0}
        \label{discrete_trajopt},
    \end{mini}
    where $x$ and $u$ are the state and control vectors as described in Sec. \ref{sec:rigidbody_dynamics},
    $f$ are the dynamics as defined in \eqref{eq:rigid_body_dynamics}, $\ell$ is a general
    nonlinear cost function,
    $N$ is the number of time steps, and $g_k$, $h_k$ are general nonlinear inequality and 
    equality constraints.

    Like most gradient or Newton-based methods for optimization, ALTRO approximates the 
    nonlinear functions $f, \ell, g,$ and $h$ with their first or second-order Taylor series
    expansions. Leveraging the methods from Sec: \ref{sec:Quaternion_Calculus}, we adapt the
    algorithm to optimize directly on the error state $\delta x \in \R^{12}$.

    We begin by linearizing the dynamics about the reference trajectory using
    \eqref{eq:quat_jacobian}. Our linearized error dynamics become
    \begin{equation} \label{eq:linearized_dynamics}
        \delta x_{k+1} = A_k \delta x_k + B_k \delta u_k 
    \end{equation}
    where \begin{equation}
        \begin{aligned}
            A_k = E(\bar{x}_{k+1})^T \pdv{f}{x}|_{\bar{x}_k,\bar{u}_k} E(\bar{x}_k), \\
            B_k = E(\bar{x}_{k+1})^T \pdv{f}{u}|_{\bar{x}_k,\bar{u}_k},
        \end{aligned}
    \end{equation}
    and $\delta x_k \in \R^{12}$ and $E(x_k) \in \R^{12 \times 13}$ are the state error and state error Jacobian:
    \begin{equation} \label{eq:state_error}
        \setlength\arraycolsep{1pt}
        \delta x_k = \begin{bmatrix} 
            r_k - \bar{r}_k \\ \varphi^{-1}(\bar{\q}_k^{-1} \otimes \q_k) \\ v_k - \bar{v}_k \\ \omega_k - \bar{\omega}_k 
        \end{bmatrix}\!, \;
        E(x) = \begin{bmatrix}
            I_3 & & & \\
            & G(q) & & \\
            & & I_3 & \\
            & & & I_3 \\
        \end{bmatrix}\!.
    \end{equation}

    By applying \eqref{eq:quat_gradient} and \eqref{eq:quat_hessian} to our nonlinear cost
    functions $\ell$ and 
    \eqref{eq:quat_jacobian} to the nonlinear constraint functions $g_k$ and $h_k$, we can calculate
    the second-order expansion 
    \begin{multline}
        \delta \ell(x,u) \approx  \half \delta x^T \ell_xx \delta x
            + \half \delta u^T \ell_{uu} \delta u + \delta_u^T \ell_{ux} \delta u \\
            + \ell_x^T \delta x^T + \ell_u^T \delta u 
    \end{multline}
    of the augmented Lagrangian cost function:
    \begin{equation}
        \mathcal{L}_A = \mathcal{L}_N(x_N,\lambda_N,\mu_N) + 
            \sum_{k=0}^{N-1} \mathcal{L}_k(x_k,u_k,\lambda_k,\mu_k)
    \end{equation}
    where
    \begin{equation}
        \mathcal{L}_k(x,u,\lambda,\mu) = \ell(x,u) + 
            (\lambda + \half I_\mu c(x,u))^T c(x,u),
    \end{equation}
    with $c(x,u)$ being the concatenation of the constraints $f,g$, and $h$ at a given time step,
    and $I_\mu$ the penalty matrix.

    With this expansion, we calculate the expansion of the ``action-value function'' $Q(x,u)$
    as normal:
    \begin{align}
        Q_{xx} &= \ell_{xx} + A_{k}^T P_{k+1} A_{k} \label{Qxx_exp}\\
        Q_{uu} &= \ell_{uu} + B_{k}^T P_{k+1} B_{k} \label{Quu_exp}\\
        Q_{ux} &= \ell_{ux} + B_{k}^T P_{k+1} A_{k} \label{Qux_exp}\\
        Q_x &= \ell_x  + A_{k}^T p_{k+1} \label{Qx_exp}\\
        Q_u &= \ell_u + B_{k}^T p_{k+1} \label{Qu_exp},
    \end{align}
    from which we can calculate the quadratic expansion of the cost-to-go 
    $P_k \in \R^{12 \times 12}$, $p_k \in \R^12$, and optimal linearized feedback gains 
    $K_k \in \R^{m \times 12}$ and $d_k \in \R^m$ by starting at the terminal state and
    resursing backward in time along the trajectory during the ``backward pass'' of the iLQR
    algorithm. During the ``forward pass'', the dynamics are simulated forward in time using
    the feedback gains computed during the backward pass. At each time step, the control is 
    calculated using the linear feedback controller: 
    \begin{equation} \label{eq:mlqr_control}
        u_k = K_k \delta x_k + \bar{u}_k.
    \end{equation}
    where $\bar{u}_k$ is the control value from the previous iteration, and $\delta x$ is 
    computed using \eqref{eq:state_error}, with $x_k$ the current state estimate and $\bar{x}_k$
    the state from the previous iteration. The rest of the algorithm is left unchanged. 
    For more details on the ALTRO algorithm, the reader is encouraged to refer to the original
    paper \cite{howell2019altro}.

    \subsection{Quaternion Cost Functions} \label{sec:cost_functions}
        In addition to the straight-forward modifications to the ALTRO algorithm itself, we
        need to carefully consider the types of cost functions we minimize. We frequently
        minimize costs that penalize distance from a goal state, e.g. $\half (x-x_g)^T Q
        (x-x_g)$; however, na\"ive substraction of unit quaternions is ill-defined. We
        propose two different cost functions that accomplish similar behavior.
        For sake of clarity and space, we only consider the costs on the quaternion
        variables: costs on the other states and the control variables remain unaffected.

@@

@@

            \begin{figure}[t]
                \centering
                \includegraphics[height=5.2cm,trim={0 3cm 0 0},clip]{figures/quad_flip.png}
                \caption{Snapshots of the quadrotor flip trajectory. The
                    red-colored quadrotors represent the state near t=0 s and the
                    green-colored quadrotors represent the state near t=5.0 s
                }
                \label{fig:quad_flip}
            \end{figure}

@@
